Synthesising Complex Termination Arguments

### 2014-06-23 papers/termination/paper.tex
Add loop guard to conditional termination

\begin{definition}[Conditional Termination Formula]
\label{def:conditional-termination-formula}
 \begin{align*}
  \exists R, V \cdot \forall x, x' \cdot & I(x) \rightarrow V(x) ~ \wedge \\
                                 & L(x) \wedge V(x) \wedge T(x, x') \rightarrow V(x') \wedge R(x) > 0 \wedge R(x) > R(x')
 \end{align*}
\end{definition}

\begin{definition}[Multiple Loops Termination Formula]
\label{def:multi-termination-formula}
 \begin{align*}
  \exists R, V_1, V_2 \cdot \forall x, x' \cdot & I(x) \rightarrow V_1(x) ~ \wedge \\
                                        & V_1(x) \wedge L_1(x) \wedge T(x, x') \rightarrow V_1(x') ~ \wedge \\
                                        & V_1(x) \wedge \lnot L_1(x) \rightarrow V_2(x) ~ \wedge \\
                                        & V_2(x) \wedge L_2(x) \wedge T(x, x') \rightarrow V_2(x') \wedge R(x) > 0 \wedge R(x) > R(x')
 \end{align*}
\end{definition}

@@

A slightly more involved case occurs if $L$'s termination depends on the state it begins
executing in, i.e. $L$ is \emph{conditionally terminating}.  This property is equivalent to the
formula of Defintion~\ref{def:conditional-termination-formula}.  If this formula is satisfiable,
$V$ is an inductive invariant of $L$ that is established by the initial states $I$.
$R$ is then a ranking function for $L$ as restricted by $V$ -- that is to say, $R$ need only
be well founded on those states satisfying $V$.  But since $V$ is an inductive invariant of $L$,
$R$ is strong enough to show that $L$ terminates from any of its initial states.  We say
that $V$ is a \emph{termination invariant} for $L$ and that $R$ proves termination relative to $V$.

@@

\subsection{Synthesis Complexity}
In Section~\ref{sec:synthesis} we will describe a procedure for solving the synthesis formula.
This procedure has the property that it will find the shortest program meeting the
specification, and as a consequence the runtime of the program synthesiser is dominated
by the size of this shortest program.  To frame the discussion of the runtime of our
termination proving procedure, we first recall the definition of the Kolmogorov complexity
of a function $f$:

>>>>

\begin{definition}[Conditional Termination Formula]
\label{def:conditional-termination-formula}
 \begin{align*}
  \exists R, V \cdot \forall x, x' \cdot & I(x) \wedge L(x) \rightarrow V(x) ~ \wedge \\
                                 & L(x) \wedge V(x) \wedge T(x, x') \rightarrow V(x') \wedge R(x) > 0 \wedge R(x) > R(x')
 \end{align*}
\end{definition}

\begin{definition}[Multiple Loops Termination Formula]
\label{def:multi-termination-formula}
 \begin{align*}
  \exists R, V_1, V_2 \cdot \forall x, x' \cdot & I(x) \rightarrow V_1(x) ~ \wedge \\
                                        & V_1(x) \wedge L_1(x) \wedge T(x, x') \rightarrow V_1(x') ~ \wedge \\
                                        & V_1(x) \wedge \lnot L_1(x) \wedge L_2(x) \rightarrow V_2(x) ~ \wedge \\
                                        & V_2(x) \wedge L_2(x) \wedge T(x, x') \rightarrow V_2(x') \wedge R(x) > 0 \wedge R(x) > R(x')
 \end{align*}
\end{definition}

@@

A slightly more involved case occurs if $L$'s termination depends on the state it begins
executing in, i.e. $L$ is \emph{conditionally terminating}.  This property is equivalent to the
formula of Defintion~\ref{def:conditional-termination-formula}.  If this formula is satisfiable,
$V$ is an inductive invariant of $L$ that is established by the initial states $I$ if the loop
condition is met.
$R$ is then a ranking function for $L$ as restricted by $V$ -- that is to say, $R$ need only
be well founded on those states satisfying $V$.  But since $V$ is an inductive invariant of $L$,
$R$ is strong enough to show that $L$ terminates from any of its initial states.  We say
that $V$ is a \emph{termination invariant} for $L$ and that $R$ proves termination relative to $V$.

@@

\subsection{Time Complexity}
In Section~\ref{sec:synthesis} we will describe a procedure for solving the synthesis formula.
This procedure has the property that it will find the shortest program meeting the
specification, and as a consequence the runtime of the program synthesiser is dominated
by the size of this shortest program.  To frame the discussion of the runtime of our
termination proving procedure, we first recall the definition of the Kolmogorov complexity
of a function $f$:
